Priors for Latent Hawkes

### 2020-06-23 output_scripts/New_Priors.Rmd
investigate the degree correction

library(rstan)
```

```{r mice_setup,include=FALSE,echo=FALSE}
cohort_names <- paste("cohort",c(9,10,12,15,16,17,18,37,38,45),sep='')
cohort_short_names <- paste("C",c(9,10,12,15,16,17,18,37,38,45),sep='')
cut_off <- 3
mice_number <- 12
```

# First Prior

To remove some of the issues with the previously specified priors,
we reparameterise `lambda1` and choose more appropriate priors
for `w_lambda,lambda0` and `beta`.

```{r eval=FALSE}

lambda1 = lambda0*(1+w_lambda);
// now w_lambda only has to be positive

//priors
lambda0 ~ gamma(0.1,1); // prior has mean 0.1, var 0.1
w_lambda ~ lognormal(0,2);
eta_1 ~ lognormal(0,1);
eta_2 ~ lognormal(0,1);
eta_3 ~ lognormal(0,1);
beta ~ lognormal(0,2);

@@

## Examine `stan` diagnostics

```{r}
fit_cohorts <- 1:10
fit_cohorts <- fit_cohorts[-7]
for(current_cohort in fit_cohorts){
  print(current_cohort)
  load(paste(data_path,cohort_names[current_cohort],
           "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
           ".RData",sep=''))
  rstan::check_hmc_diagnostics(fit_cohort_mmhp)
}

This leads to better results in terms of divergent transitions and
max tree depth, which were both issues previously.

## Param Histograms

We can check the traceplots/posterior distributions
based on these prior choices.

```{r}
current_cohort <- 1
print(current_cohort)
load(paste(data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
         ".RData",sep=''))
rstan::stan_hist(fit_cohort_mmhp,
               pars = c("lambda0","lambda1","w_lambda",
                        "eta_1","eta_2","eta_3","beta")) +
  ggtitle("New Priors")
rstan::traceplot(fit_cohort_mmhp,
               pars = c("lambda0","lambda1","w_lambda",
                        "eta_1","eta_2","eta_3","beta")) + 
  ggtitle("New Priors")

We can similarly look at the original fits (with and without considering the windows with 
no events).

```{r}

new_data_path <- "C:/Users/owenw/Google Drive/Tian/Current_Projects/MMHP_Latent/debug_copy/"

load(paste(orig_data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
         ".RData",sep=''))

rstan::stan_hist(fit_cohort_mmhp,
               pars = c("lambda0","lambda1","w_lambda",
                        "eta_1","eta_2","eta_3","beta")) +
  ggtitle("Original, no null windows")
rstan::traceplot(fit_cohort_mmhp,
               pars = c("lambda0","lambda1","w_lambda",
                        "eta_1","eta_2","eta_3","beta")) + 
  ggtitle("Original, no null windows")

fit_cohort_mmhp <- readRDS(paste(new_data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_fit_",cohort_names[current_cohort],
         ".RDS",sep=''))

rstan::stan_hist(fit_cohort_mmhp,
               pars = c("lambda0","lambda1","w_lambda",
                        "eta_1","eta_2","eta_3","beta")) +
  ggtitle("Original, null windows")
rstan::traceplot(fit_cohort_mmhp,
               pars = c("lambda0","lambda1","w_lambda",
                        "eta_1","eta_2","eta_3","beta")) + 
  ggtitle("Original, null windows")

## Latent Rankings

```{r}
load(paste(data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
         ".RData",sep=''))

@@

We can compare this to the original priors (both with and excluding the null windows).

```{r}
load(paste(orig_data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
         ".RData",sep=''))
p2 <- rstan::traceplot(fit_cohort_mmhp,
                 pars = c("f")) + 
  ggtitle("Original, no null windows")

fit_cohort_mmhp <- readRDS(paste(new_data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_fit_",cohort_names[current_cohort],
         ".RDS",sep=''))

p3 <- rstan::traceplot(fit_cohort_mmhp,
                 pars = c("f")) + 
  ggtitle("Original, null windows")

p2
p3

```{r,fig.height=9}
ggpubr::ggarrange(p1,p2,p3,nrow=3)
```

These seem similar either way, and maybe tighter with these new priors.

@@

We run into some issues with this where `w_lambda` can explode,
likely caused by extremely small draws for `lambda0`. To circumvent this
we place a lower bound on `lambda0` of 0.01 and refit this model.

@@

This avoids the previous issue with parameters blowing up however one chain
seems to be stuck.

```{r}
  print(current_cohort)
  load(paste(data_path,cohort_names[current_cohort],
           "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
           ".RData",sep=''))
  rstan::check_hmc_diagnostics(fit_cohort_mmhp)
```
```{r}
p1_new <-rstan::traceplot(fit_cohort_mmhp,
                 pars = c("f")) + 
  ggtitle("New Prior, lower bound on lambda")

@@

We can compare this to the previous fit.

@@

@@

>>>>

func_dir <- "../lib/"

library(rstan)
Sys.setenv(LOCAL_CPPFLAGS = '-march=corei7 -mtune=corei7')
options(mc.cores = parallel::detectCores())
library(tidyverse)
library(R.utils)
library(fields)
library(bayesplot)
library(compete)
source(paste(func_dir,'naiveRankHierarchy.R',sep = ""))
source(paste(func_dir,'expertRankHierarchy.R',sep = ""))
source(paste(func_dir,'cleanData.R',sep = "")) 
source(paste(func_dir,'prepareDataStan.R',sep = "")) 
source(paste(func_dir,'inferLatentMmhp.R',sep = ""))
source(paste(func_dir,'plotUtil.R',sep = "")) 
source(paste(func_dir,'mmhp.R',sep = ""))
source(paste(func_dir,'uniHawkes.R',sep = ""))
source(paste(func_dir,'simulatePrediction.R',sep = ""))
source(paste(func_dir,'myGlicko.R',sep = ""))
source(paste(func_dir,'matrixPlotParameter.R',sep = ""))
source(paste(func_dir,'residualStructureScore.R',sep = ""))
source("https://gist.githubusercontent.com/jalapic/6ca3ece44bdcdc522bb735f183aa0ca0/raw/1a07f469eff08117121b6cbebcd22fb7569e3ee8/compete_extra.R")

theme_set(theme_minimal())
```

```{r mice_setup,include=FALSE,echo=FALSE}
cohort_names <- paste("cohort",c(9,10,12,15,16,17,18,37,38,45),sep='')
cohort_short_names <- paste("C",c(9,10,12,15,16,17,18,37,38,45),sep='')
cut_off <- 3
mice_number <- 12

full_data <- readRDS("../data/mice.RData")
# A=c9, B=c10, C=c12, D=c15, E=c16, F=c17, G=c18, H=c37, I=c38. J=c45
cohort_names <- paste("cohort",c(9,10,12,15,16,17,18,37,38,45),sep='')
cohort_short_names <- paste("C",c(9,10,12,15,16,17,18,37,38,45),sep='')
cut_off <- 3
mice_number <- 12

# Define the cohorts will be fitted
fit_cohorts <- c(1:10)
naive_rank_10 <- list()
expert_rank_10 <- list()
for(current_cohort in fit_cohorts){
  naive_rank_10[[current_cohort]] <- naiveRankHierarchy(full_data[[cohort_names[current_cohort]]])
  expert_rank_10[[current_cohort]] <- expertRankHierarchy(full_data[[cohort_names[current_cohort]]])
}

@@

<!-- # First Prior -->

<!-- To remove some of the issues with the previously specified priors, -->
<!-- we reparameterise `lambda1` and choose more appropriate priors -->
<!-- for `w_lambda,lambda0` and `beta`. -->

<!-- ```{r eval=FALSE} -->

<!-- lambda1 = lambda0*(1+w_lambda); -->
<!-- // now w_lambda only has to be positive -->

<!-- //priors -->
<!-- lambda0 ~ gamma(0.1,1); // prior has mean 0.1, var 0.1 -->
<!-- w_lambda ~ lognormal(0,2); -->
<!-- eta_1 ~ lognormal(0,1); -->
<!-- eta_2 ~ lognormal(0,1); -->
<!-- eta_3 ~ lognormal(0,1); -->
<!-- beta ~ lognormal(0,2); -->

<!-- ``` -->

<!-- ## Examine `stan` diagnostics -->

<!-- ```{r} -->
<!-- fit_cohorts <- 1:10 -->
<!-- fit_cohorts <- fit_cohorts[-7] -->
<!-- for(current_cohort in fit_cohorts){ -->
<!--   print(current_cohort) -->
<!--   load(paste(data_path,cohort_names[current_cohort], -->
<!--            "/cohort_mmhp_stan_result_",cohort_names[current_cohort], -->
<!--            ".RData",sep='')) -->
<!--   rstan::check_hmc_diagnostics(fit_cohort_mmhp) -->
<!-- } -->

<!-- ``` -->

<!-- This leads to better results in terms of divergent transitions and -->
<!-- max tree depth, which were both issues previously. -->

<!-- ## Param Histograms -->

<!-- We can check the traceplots/posterior distributions -->
<!-- based on these prior choices. -->

<!-- ```{r} -->
<!-- current_cohort <- 1 -->
<!-- print(current_cohort) -->
<!-- load(paste(data_path,cohort_names[current_cohort], -->
<!--          "/cohort_mmhp_stan_result_",cohort_names[current_cohort], -->
<!--          ".RData",sep='')) -->
<!-- rstan::stan_hist(fit_cohort_mmhp, -->
<!--                pars = c("lambda0","lambda1","w_lambda", -->
<!--                         "eta_1","eta_2","eta_3","beta")) + -->
<!--   ggtitle("New Priors") -->
<!-- rstan::traceplot(fit_cohort_mmhp, -->
<!--                pars = c("lambda0","lambda1","w_lambda", -->
<!--                         "eta_1","eta_2","eta_3","beta")) +  -->
<!--   ggtitle("New Priors") -->

<!-- ``` -->

<!-- We can similarly look at the original fits (with and without considering the windows with  -->
<!-- no events). -->

<!-- ```{r} -->

<!-- orig_data_path <- "C:/Users/owenw/Google Drive/Tian/Research+JingWu/thesis_data/part2/real_data/" -->

<!-- new_data_path <- "C:/Users/owenw/Google Drive/Tian/Current_Projects/MMHP_Latent/debug_copy/" -->

<!-- load(paste(orig_data_path,cohort_names[current_cohort], -->
<!--          "/cohort_mmhp_stan_result_",cohort_names[current_cohort], -->
<!--          ".RData",sep='')) -->

<!-- rstan::stan_hist(fit_cohort_mmhp, -->
<!--                pars = c("lambda0","lambda1","w_lambda", -->
<!--                         "eta_1","eta_2","eta_3","beta")) + -->
<!--   ggtitle("Original, no null windows") -->
<!-- rstan::traceplot(fit_cohort_mmhp, -->
<!--                pars = c("lambda0","lambda1","w_lambda", -->
<!--                         "eta_1","eta_2","eta_3","beta")) +  -->
<!--   ggtitle("Original, no null windows") -->

<!-- fit_cohort_mmhp <- readRDS(paste(new_data_path,cohort_names[current_cohort], -->
<!--          "/cohort_mmhp_stan_fit_",cohort_names[current_cohort], -->
<!--          ".RDS",sep='')) -->

<!-- rstan::stan_hist(fit_cohort_mmhp, -->
<!--                pars = c("lambda0","lambda1","w_lambda", -->
<!--                         "eta_1","eta_2","eta_3","beta")) + -->
<!--   ggtitle("Original, null windows") -->
<!-- rstan::traceplot(fit_cohort_mmhp, -->
<!--                pars = c("lambda0","lambda1","w_lambda", -->
<!--                         "eta_1","eta_2","eta_3","beta")) +  -->
<!--   ggtitle("Original, null windows") -->

<!-- ``` -->

<!-- ## Latent Rankings -->

```{r, include=FALSE}
current_cohort <- 1
load(paste(data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
         ".RData",sep=''))

@@

<!-- We can compare this to the original priors (both with and excluding the null windows). -->

<!-- ```{r} -->
<!-- load(paste(orig_data_path,cohort_names[current_cohort], -->
<!--          "/cohort_mmhp_stan_result_",cohort_names[current_cohort], -->
<!--          ".RData",sep='')) -->
<!-- p2 <- rstan::traceplot(fit_cohort_mmhp, -->
<!--                  pars = c("f")) +  -->
<!--   ggtitle("Original, no null windows") -->

<!-- fit_cohort_mmhp <- readRDS(paste(new_data_path,cohort_names[current_cohort], -->
<!--          "/cohort_mmhp_stan_fit_",cohort_names[current_cohort], -->
<!--          ".RDS",sep='')) -->

<!-- p3 <- rstan::traceplot(fit_cohort_mmhp, -->
<!--                  pars = c("f")) +  -->
<!--   ggtitle("Original, null windows") -->

<!-- p2 -->
<!-- p3 -->

<!-- ``` -->

<!-- ```{r,fig.height=9} -->
<!-- ggpubr::ggarrange(p1,p2,p3,nrow=3) -->
<!-- ``` -->

<!-- These seem similar either way, and maybe tighter with these new priors. -->

@@

We have modified the current priors so that 
$\lambda_1 = \lambda_0(1+w_{\lambda})$ with lognormal priors for
all parameters. Now $w_{\lambda}>0$ as opposed to $>1$ previously.

We ran into some issues with this where `w_lambda` can explode,
likely caused by extremely small draws for `lambda0`. To circumvent this
we place a lower bound on `lambda0` of 0.01 and refit this model.

The resulting posterior draws are shown below for a single cohort.

@@

This avoids the previous issue with parameters blowing up however one chain
seems to be stuck. This doesn't occur for other cohorts, and is likely not
a problem.

```{r}
  print(current_cohort)
  load(paste(data_path,cohort_names[current_cohort],
           "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
           ".RData",sep=''))
  rstan::check_hmc_diagnostics(fit_cohort_mmhp)
```

Similarly, we can look at the inferred latent ranks.

```{r}
p1_new <-rstan::traceplot(fit_cohort_mmhp,
                 pars = c("f")) + 
  ggtitle("New Prior, lower bound on lambda")

@@

We can compare this to the previous fit (without this lower bound).

@@

We instead fit a degree corrected C-MMHP. Here we have

$$
\lambda_{0}^{i,j} = \gamma_i + \zeta_j
$$

for each pair. We use Laplacian priors on these (used in DC-HP). We
can explore the resulting fits.

This seems to avoid the issue with $w_{\lambda}$ taking extremely large values
also.

@@

## Compare posterior ranks

Here we plot the latent ranks in the previous section against those
obtained with the degree correction.

```{r, fig.height=10}
current_cohort <- 1
print(current_cohort)
data_path <- "C:/Users/owenw/Google Drive/Tian/Current_Projects/MMHP_Latent/output_p1//"
load(paste(data_path,cohort_names[current_cohort],
           "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
           ".RData",sep=''))
p1 <-rstan::traceplot(fit_cohort_mmhp,
                 pars = c("f")) + 
  ggtitle("Bound lambda0")

data_path <- "C:/Users/owenw/Google Drive/Tian/Current_Projects/MMHP_Latent/output_dc_prior//"
load(paste(data_path,cohort_names[current_cohort],
           "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
           ".RData",sep=''))
p1_new <-rstan::traceplot(fit_cohort_mmhp,
                 pars = c("f")) + 
  ggtitle("Degree Corrected")

load(paste(orig_data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
         ".RData",sep=''))
p2 <- rstan::traceplot(fit_cohort_mmhp,
                 pars = c("f")) +
  ggtitle("Original, no null windows")

ggpubr::ggarrange(p1,p1_new,p2,nrow=3)
```

## Plot ranks against degree parameters

As before, we can also plot these inferred degree corrected rates against
the I&SI rankings.

```{r}
cohorts <- 1:10
#cohorts <- cohorts[c(-3,-6,-7)]
for(current_cohort in cohorts){
  #cat(paste("Cohort ",current_cohort,sep = ""))
  load(paste(dc_data_path,cohort_names[current_cohort],
         "/cohort_mmhp_stan_result_",cohort_names[current_cohort],
         ".RData",sep=''))

  mean_rank <- apply(sim_cohort_mmhp$f,2,mean)

  mean_gamma <- apply(sim_cohort_mmhp$gamma,2,mean)
  mean_zeta <- apply(sim_cohort_mmhp$zeta, 2, mean)

  mean_values <- tibble(mean_rank,mean_gamma,mean_zeta,
                        isi = expert_rank_10[[current_cohort]])
  p1 <- mean_values %>%
    rename(gamma_out = mean_gamma,zeta_in = mean_zeta) %>%
    pivot_longer(cols = c(gamma_out,zeta_in),names_to = "out_in" ) %>%
    mutate(out_in = as.factor(out_in)) %>%
    ggplot(aes(isi,value))  + geom_point()  +
    theme(legend.title = element_blank()) + facet_wrap(~out_in,scales="free") +
    scale_x_continuous(breaks = 1:12) + xlab("I&SI Ranking") +
    ggtitle(paste("Cohort",current_cohort,sep=""))
  print(p1)

}
```

This seems to indicate that the degree correction is somewhat useful for 
certain mice in providing a more expressive fit.

I'm running this through for the rest of the inference (predictions,
redisuals) at the moment.
